---
title: "Final Project"
author: "Elizabeth McHugh"
date: "5/22/2021"
output:
  pdf_document:
    toc: yes
    fig_caption: yes
    keep_tex: yes
  word_document:
    toc: yes
---

```{r}

if (!require("pacman")){install.packages("pacman")}
pacman::p_load(knitr, randomForest, dplyr, tidyverse, ggplot2, missForest, stats, readr, magrittr, data.table, lubridate, stringr)

```

##Import and Clean Data Set (2.1)##

*Import the data set from drive.*

```{r}

library(readr)
housing_data_2016_2017 <- read_csv("C:\\Users\\twiz0\\Downloads\\housing_data_2016_2017.csv")
#View(housing_data_2016_2017)

```

#Split Data#

Split Test and Training Sets. Retain 20% of data for testing.

```{r}

set.seed(479)

#Split 20% Test/ 80% Train
K = 5   

test_indices = sample(1 : nrow(housing_data_2016_2017), round(nrow(housing_data_2016_2017) / K))
train_indices = setdiff(1 : nrow(housing_data_2016_2017), test_indices)

housing_data_test = housing_data_2016_2017[test_indices, ]
housing_data_train = housing_data_2016_2017[train_indices, ]

#View(housing_data_train)
#View(housing_data_test)

#summary(housing_data_train)
#summary(housing_data_test)

#Count observations with missing target variable.
sum(is.na(housing_data_2016_2017$sale_price))
```

#Initial (Pre-Imputation) Data Clean-up (2.2)#

Data Clean-Up on Training Set

```{r}

#Remove obviously unnecessary columns, reorder with objective variable (sale price) at head, remove observations with missing sale price. 
housing_data_train = housing_data_train %>%
  select(-(1:28), -url) %>%
    select(sale_price, everything()) %>%
      filter(!is.na(sale_price)) %>%
        select(-listing_price_to_nearest_1000)

#Unformat all prices
housing_data_train = housing_data_train %>%
  mutate(sale_price = parse_number(sale_price)) %>%
  mutate(common_charges = parse_number(common_charges)) %>%
  mutate(maintenance_cost = parse_number(maintenance_cost)) %>%
  mutate(parking_charges = parse_number(parking_charges)) %>%
  mutate(total_taxes = parse_number(total_taxes))

#Add feature for total bathrooms (whole plus half). 
housing_data_train = housing_data_train %>%
  mutate(num_half_bathrooms = replace(num_half_bathrooms, is.na(num_half_bathrooms), 0)) %>%
  mutate(num_bathrooms = num_full_bathrooms + 0.5 * num_half_bathrooms) 

#Separate dates sold as year, date, month, weekdays, and days of month.
housing_data_train = housing_data_train %>%
  mutate(date_of_sale = as_date(mdy(date_of_sale))) %>%
  mutate(month_of_year = month(date_of_sale)) %>%
  mutate(day_of_week = wday(date_of_sale)) %>%
  mutate(day_of_month = as.numeric(day(date_of_sale))) %>%
  mutate(year = year(date_of_sale)) %>%
    mutate(date_of_sale = as.numeric(date_of_sale))

#Extract zip codes from addresses.
housing_data_train = housing_data_train %>%
  mutate(zip_numeric = as.numeric(str_sub(full_address_or_zip_code, -5,-1))) %>%
  mutate(zip_factor = as.factor(zip_numeric))

#Create dummy variables for non-factor variables with potentially significant missing data.
housing_data_train = housing_data_train %>%
  mutate(common_charges_missing = as.factor(is.na(common_charges))) %>%
    mutate(common_charges = ifelse(is.na(common_charges), 0, common_charges)) %>%
  mutate(approx_year_built_missing = as.factor(is.na(approx_year_built))) %>%
  mutate(maintenance_cost_missing = as.factor(is.na(maintenance_cost))) %>%
    mutate(maintenance_cost = ifelse(is.na(maintenance_cost), 0, maintenance_cost)) %>%
  mutate(num_floors_in_building_missing = as.factor(is.na(num_floors_in_building))) %>%
  mutate(parking_charges_missing = as.factor(is.na(parking_charges))) %>%
    mutate(parking_charges = ifelse(is.na(parking_charges), 0, parking_charges)) %>%
  mutate(pct_tax_deductibl_missing = as.factor(is.na(pct_tax_deductibl))) %>%
  mutate(sq_footage_missing = as.factor(is.na(sq_footage))) %>%
  mutate(total_taxes_missing = as.factor(is.na(total_taxes)))

#Coerce yes/no to factors.
housing_data_train = housing_data_train %>%
  mutate(cats_allowed = factor(cats_allowed)) %>%
  mutate(dogs_allowed = factor(dogs_allowed))

#Garage exists to factor.
housing_data_train = housing_data_train %>%
  mutate(garage_exists = as.factor(!is.na(garage_exists))) 

#Factorize character variables and set NA to "unknown" factor.
housing_data_train = housing_data_train %>%
  mutate(dining_room_type = replace_na(dining_room_type, "unknown")) %>%
  mutate(dining_room_type = factor(dining_room_type)) %>%
  mutate(coop_condo = factor(coop_condo, ordered = FALSE)) %>%
  mutate(fuel_type = ifelse(fuel_type %in% c("other", "Other"), "other", fuel_type)) %>%
  mutate(fuel_type = ifelse(is.na(fuel_type), "unknown", fuel_type)) %>%
  mutate(fuel_type = factor(fuel_type)) %>%
  mutate(kitchen_type = ifelse(kitchen_type %in% c("eat in", "Eat In", "Eat in"), "eat in", kitchen_type)) %>%
  mutate(kitchen_type = replace_na(kitchen_type, "unknown")) %>%
  mutate(kitchen_type = ifelse(kitchen_type == "Combo", "combo", kitchen_type)) %>% 
  mutate(kitchen_type = as.factor(kitchen_type))

#Take care of factors with only a few observations.
housing_data_train = housing_data_train %>%
  mutate(dining_room_type = recode(dining_room_type, "dining area" = "other")) %>%
  mutate(kitchen_type = recode(kitchen_type, "1955" = "unknown"))

#Fill in singular missing values in train data (found zip manually in raw data)
housing_data_train$zip_numeric[2] = 11354
housing_data_train$zip_factor[2] = "11354"

#Remove full address, model type, and date of sale
housing_data_train = housing_data_train %>%
  mutate(total_additional_charges = common_charges + maintenance_cost + parking_charges) %>%
  select(-full_address_or_zip_code, -model_type, -common_charges, -parking_charges, -maintenance_cost)
  
View(summary(housing_data_train))
#sapply(housing_data_train, class)
```

Data Clean-Up on Test Set

```{r}

#Remove obviously unnecessary columns, reorder with objective variable (sale price) at head, remove observations with missing sale price. 
housing_data_test = housing_data_test %>%
  select(-(1:28), -url) %>%
    select(sale_price, everything()) %>%
      filter(!is.na(sale_price)) %>%
        select(-listing_price_to_nearest_1000)

#Unformat all prices
housing_data_test = housing_data_test %>%
  mutate(sale_price = parse_number(sale_price)) %>%
  mutate(common_charges = parse_number(common_charges)) %>%
  mutate(maintenance_cost = parse_number(maintenance_cost)) %>%
  mutate(parking_charges = parse_number(parking_charges)) %>%
  mutate(total_taxes = parse_number(total_taxes))

#Add feature for total bathrooms (whole plus half). 
housing_data_test = housing_data_test %>%
  mutate(num_half_bathrooms = replace(num_half_bathrooms, is.na(num_half_bathrooms), 0)) %>%
  mutate(num_bathrooms = num_full_bathrooms + 0.5 * num_half_bathrooms) 

#Separate dates sold as year, date, month, weekdays, and days of month.
housing_data_test = housing_data_test %>%
  mutate(date_of_sale = as_date(mdy(date_of_sale))) %>%
  mutate(month_of_year = month(date_of_sale)) %>%
  mutate(day_of_week = wday(date_of_sale)) %>%
  mutate(day_of_month = as.numeric(day(date_of_sale))) %>%
  mutate(year = year(date_of_sale)) %>%
    mutate(date_of_sale = as.numeric(date_of_sale))

#Extract zip codes from addresses.
housing_data_test = housing_data_test %>%
  mutate(zip_numeric = as.numeric(str_sub(full_address_or_zip_code, -5,-1))) %>%
  mutate(zip_factor = as.factor(zip_numeric))

#Create dummy variables for non-factor variables with potentially significant missing data.
housing_data_test = housing_data_test %>%
  mutate(common_charges_missing = as.factor(is.na(common_charges))) %>%
    mutate(common_charges = ifelse(is.na(common_charges), 0, common_charges)) %>%
  mutate(approx_year_built_missing = as.factor(is.na(approx_year_built))) %>%
  mutate(maintenance_cost_missing = as.factor(is.na(maintenance_cost))) %>%
    mutate(maintenance_cost = ifelse(is.na(maintenance_cost), 0, maintenance_cost)) %>%
  mutate(num_floors_in_building_missing = as.factor(is.na(num_floors_in_building))) %>%
  mutate(parking_charges_missing = as.factor(is.na(parking_charges))) %>%
    mutate(parking_charges = ifelse(is.na(parking_charges), 0, parking_charges)) %>%
  mutate(pct_tax_deductibl_missing = as.factor(is.na(pct_tax_deductibl))) %>%
  mutate(sq_footage_missing = as.factor(is.na(sq_footage))) %>%
  mutate(total_taxes_missing = as.factor(is.na(total_taxes)))

#Coerce yes/no to factors.
housing_data_test = housing_data_test %>%
  mutate(cats_allowed = factor(cats_allowed)) %>%
  mutate(dogs_allowed = factor(dogs_allowed))

#Garage exists to factor.
housing_data_test = housing_data_test %>%
  mutate(garage_exists = as.factor(!is.na(garage_exists))) 

#Factorize character variables and set NA to "unknown" factor.
housing_data_test = housing_data_test %>%
  mutate(dining_room_type = replace_na(dining_room_type, "unknown")) %>%
  mutate(dining_room_type = factor(dining_room_type)) %>%
  mutate(coop_condo = factor(coop_condo, ordered = FALSE)) %>%
  mutate(fuel_type = ifelse(fuel_type %in% c("other", "Other"), "other", fuel_type)) %>%
  mutate(fuel_type = ifelse(is.na(fuel_type), "unknown", fuel_type)) %>%
  mutate(fuel_type = factor(fuel_type)) %>%
  mutate(kitchen_type = ifelse(kitchen_type %in% c("eat in", "Eat In", "Eat in"), "eat in", kitchen_type)) %>%
  mutate(kitchen_type = replace_na(kitchen_type, "unknown")) %>%
  mutate(kitchen_type = ifelse(kitchen_type == "Combo", "combo", kitchen_type)) %>% 
  mutate(kitchen_type = as.factor(kitchen_type))

#Take care of factors with only a few observations.
housing_data_test = housing_data_test %>%
  mutate(dining_room_type = recode(dining_room_type, "dining area" = "other")) %>%
  mutate(kitchen_type = recode(kitchen_type, "1955" = "unknown"))

#Fill in singular missing values easily available manually 
housing_data_test = housing_data_test %>%
  mutate(dining_room_type = recode(dining_room_type, "dining area" = "other"))

#Remove full address, model type, and date of sale
housing_data_test = housing_data_test %>%
  mutate(total_additional_charges = common_charges + maintenance_cost + parking_charges) %>%
  select(-full_address_or_zip_code, -model_type, -common_charges, -parking_charges, -maintenance_cost)

summary(housing_data_test)
#sapply(housing_data_test, class)
```

##Missingness in Features (2.3)

Impute using missForest.

```{r}

#Impute missing values in training data
housing_data_train_imp = missForest(data.frame(housing_data_train))$ximp

#Impute missing values in test data.
housing_data_test_imp = cbind("sale_price" = NA, housing_data_test[2:ncol(housing_data_test)])
housing_data_test_train_imp = rbind(housing_data_test_imp, housing_data_train_imp)
housing_data_test_train_imp = missForest(data.frame(housing_data_test_train_imp))$ximp

housing_data_test_imp = housing_data_test_train_imp[1:nrow(housing_data_test_imp), ]

```

Playing with visualizations to consider feature transformations.

```{r}

#Not a linear relationship.
ggplot(housing_data_train_imp) +
  aes(x = log(total_additional_charges), y = sale_price) +
  geom_smooth() +
  geom_jitter()


#Note how the relative lack of data in zip codes (just two zips?) below 11300 as well as the lack of density in sales price in those zip codes (this would seem to not make for good prediction) as compared to the apparent negative correlation between sale price and zip code above 11300 (which actually might serve as a decent predictor for sales price in those zip codes). 


ggplot(housing_data_train_imp) +
  aes(x = zip_factor, y = sale_price) +
  geom_smooth() +
  geom_jitter()

ggplot(housing_data_train_imp) +
  aes(x = zip_numeric, y = sale_price) +
  geom_smooth() +
  geom_jitter()

#Visualize effect of interactions between #bedrooms and #bathrooms on sale price
ggplot(housing_data_train_imp) +
  aes(x = (num_bedrooms / num_bathrooms)^2, y = sale_price) +
  geom_smooth() + 
  geom_jitter()

ggplot(housing_data_train_imp) +
  aes(x = (num_bedrooms / num_bathrooms)^2, y = sale_price) +
  geom_smooth() + 
  geom_jitter()

```

Feature Transformations

Add feature transformations to be included in models.

```{r}
#Training Data Transformations

housing_data_train_imp = housing_data_train_imp %>%
  mutate(log_tot_add_charges = log(total_additional_charges)) %>%
  mutate(log_tot_add_charges = ifelse(log_tot_add_charges == -Inf, 0, log_tot_add_charges)) %>%
    select(-num_half_bathrooms) %>%
    mutate(num_missing = (as.numeric(common_charges_missing) + as.numeric(approx_year_built_missing) + as.numeric(maintenance_cost_missing) + as.numeric(num_floors_in_building_missing) + as.numeric(parking_charges_missing) + as.numeric(pct_tax_deductibl_missing) + as.numeric(sq_footage_missing) + as.numeric(total_taxes_missing))) %>%
      select(-common_charges_missing, -approx_year_built_missing, -maintenance_cost_missing, -num_floors_in_building_missing, -parking_charges_missing, -pct_tax_deductibl_missing, -sq_footage_missing)

housing_data_train_imp = housing_data_train_imp %>%
  mutate(bedroom_sq_ft_ratio = num_bedrooms / sq_footage) %>%
  mutate(bedroom_bathroom_ratio = num_bedrooms / num_bathrooms) %>%
    select(-zip_numeric)

#Test Data Transformations


housing_data_test_imp = housing_data_test_imp %>%
  mutate(log_tot_add_charges = log(total_additional_charges)) %>%
  mutate(log_tot_add_charges = ifelse(log_tot_add_charges == -Inf, 0, log_tot_add_charges)) %>%
    select(-num_half_bathrooms) %>%
    mutate(num_missing = (as.numeric(common_charges_missing) + as.numeric(approx_year_built_missing) + as.numeric(maintenance_cost_missing) + as.numeric(num_floors_in_building_missing) + as.numeric(parking_charges_missing) + as.numeric(pct_tax_deductibl_missing) + as.numeric(sq_footage_missing) + as.numeric(total_taxes_missing))) %>%
      select(-common_charges_missing, -approx_year_built_missing, -maintenance_cost_missing, -num_floors_in_building_missing, -parking_charges_missing, -pct_tax_deductibl_missing, -sq_footage_missing)

housing_data_test_imp = housing_data_test_imp %>%
  mutate(bedroom_sq_ft_ratio = num_bedrooms / sq_footage) %>%
  mutate(bedroom_bathroom_ratio = num_bedrooms / num_bathrooms) %>%
    select(-zip_numeric)


#head(housing_data_train_imp)
#head(housing_data_test_imp)

```

Split into X, y test and training sets.

```{r}

X_train = housing_data_train_imp[ , 2:ncol(housing_data_train_imp)]
y_train = housing_data_train_imp[ , 1]

X_test = housing_data_test_imp[ , 2:ncol(housing_data_test_imp)]
y_test = housing_data_test[ ,1]


pacman::p_load(gtsummary)
tbl_summary(rbind(X_train, X_test), statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} / {N} ({p}%)")) 
```

##Regression Tree Modeling (3.1)

Load YARF

```{r}

Sys.setenv(JAVA_HOME = '/usr/lib/jvm/jdk1.8.0_65')

if (!pacman::p_isinstalled(YARF)){
  pacman::p_install_gh("kapelner/YARF/YARFJARs", ref = "dev")
  pacman::p_install_gh("kapelner/YARF/YARF", ref = "dev", force = TRUE)
}
options(java.parameters = "-Xmx4000m")
pacman::p_load(YARF)

library(YARF, YARFJARs)

```

Create one tree model.

```{r}

mod_YARF = YARF(y = y_train, X = X_train, num_trees = 1)

illustrate_trees(mod_YARF, max_depth = 5, length_in_px_per_half_split = 30, font_size = 14, line_rgb_color = c(0, 200, 100), background_rgb_color = c(250, 250, 250), text_rgb_color = c(50, 0, 100), file_format = "png", title = "Housing Data YARF Model Tree", open_file = TRUE)

mod_YARF

```

Tree Metrics? Nope.. Just a free space to check out some things.

```{r}

#housing_data_test
#housing_data_train_imp

```

##Linear Modeling (3.2)

Create OLS Model

```{r}

#summary(X_train)
#str(X_train)

mod_ols = lm(y_train ~ ., X_train)
mod_ols

summary(mod_ols)
View(data.frame(coefficients(mod_ols)), "OLS Model Coefficients")
View(mod_ols)
```

OLS In-Sample Metrics

```{r}

RMSE = summary(mod_ols)$sigma
RMSE

r_squared = summary(mod_ols)$r.square

View(data.frame(cbind("R Squared" = r_squared, "RMSE" = RMSE)),  title = "OLS Model In-Sample Errors")

```

##Random Forest Modeling (3.3)

Create RF Model

```{r}

rf_mod = randomForest(y_train ~ . , data = X_train, ntree = 6000, mtry = 25)

rf_mod_YARF = YARF(X = X_train, y = y_train, num_trees = 6000, mtry = 25)

```

##Performance Results for Random Forest (4)

RF Metrics

```{r}

rf_mod
rf_mod_YARF

oob_se = sd(housing_data_train$sale_price - rf_mod$predicted)
oob_se

View(data.frame(cbind("R-Squared" = max(rf_mod$rsq), "OOB_SE" = oob_se)), "Random Forest Metrics")

```

#Break open the test data.

Out-of-sample OLS model metrics

```{r}

y_test = as.matrix(y_test)

y_hat_oos = predict(mod_ols, X_test)
oos_residuals = y_test - y_hat_oos

R_sq_oos = 1 - sum(oos_residuals^2) / sum((y_test - mean(y_test))^2)
RMSE_oos = sqrt(mean(oos_residuals^2))
ooss_e = sd(y_hat_oos - y_test)

RMSE_oos
R_sq_oos
ooss_e

```

Create a final OLS model and compute final in-sample statistics for
whole data set.

```{r}

train = cbind(X_train, "sale_price" = y_train)
test = cbind(X_test, y_test)
full = rbind(train, test)

head(train)
head(test)

X = full[ , 1:(ncol(full) - 1)]
y = full[ , ncol(full)]
  
ols_mod_final = lm(y ~ ., X)
summary(ols_mod_final)

summary(ols_mod_final)$r.sq

R_sq_final = summary(ols_mod_final)$r.sq
RMSE_final = summary(ols_mod_final)$sigma

```

```{r}

RMSE_Rsq_table = data.frame(cbind("RMSE" = c(RMSE, RMSE_oos, RMSE_final), "R Squared" = c(r_squared, R_sq_oos, R_sq_final)), row.names = c("In Sample", "Out-of-Sample", "Final Model In-Sample"))

View(RMSE_Rsq_table)

```

```{r}
#OOS randomForest
y_test = as.matrix(y_test)
y_hat_rf_mod = predict(rf_mod, X_test)
oos_residuals_rf_mod = y_test - y_hat_rf_mod

R_sq_rf_mod = 1 - sum(oos_residuals_rf_mod^2) / sum((y_test - mean(y_test))^2)
RMSE_rf_mod = sqrt(mean(oos_residuals_rf_mod^2))
ooss_e_rf_mod = sd(y_hat_rf_mod - y_test)

RMSE_rf_mod
R_sq_rf_mod
ooss_e_rf_mod

#OOS YARF
y_test = as.matrix(y_test)
y_hat_YARF = predict(rf_mod_YARF, X_test)
oos_residuals_YARF = y_test - y_hat_YARF

R_sq_YARF = 1 - sum(oos_residuals_YARF^2) / sum((y_test - mean(y_test))^2)
RMSE_YARF = sqrt(mean(oos_residuals_YARF^2))
ooss_e_YARF = sd(y_hat_YARF - y_test)

RMSE_YARF
R_sq_YARF
ooss_e_YARF

data.frame(cbind(cbind("RMSE_YARF" = RMSE_YARF, "R_sq_YARF" = R_sq_YARF, "ooss_e_YARF" = ooss_e_YARF)), cbind("RMSE_rf_mod" = RMSE_rf_mod, "R_sq_rf_mod" = R_sq_rf_mod, "ooss_e_rf_mod" = ooss_e_rf_mod))
```

```{r}
#Final Model randomForest
rf_final_mod = randomForest(y ~ . , data = X, ntree = 6000, mtry = 25)
rf_final_mod

```

```{r}

#Final Model YARF
YARF_final_model = YARF(X = X_train, y = y_train, num_trees = 6000, mtry = 25)
YARF_final_model

```

```{r}

oob_se_rf_fin = sd(y - rf_final_mod$predicted)
oob_se_rf_fin

r_sq_rf_fin = max(rf_mod$rsq)
r_sq_rf_fin

oob_se_YARF_fin = sd(y - rf_final_mod$predicted)
oob_se_YARF_fin

r_sq_YARF_fin = YARF_final_model$r_squared
r_sq_YARF_fin

```
